Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-07-20 figuresANDmanuscript/manuscriptBMCformat/manuscript_rnaseq.tex
A first day of work on the MS tex file

  \section*{Results}
Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental overview}
We performed mRNA sequencing of caecum tissue of mice infected with
the parasite \textit{E. falciformis} at different timepoints and for
infections of different mouse strains.

Replicate average of transcripts as order of magnitude for
\textit{E. falciformis}/mouse are indicated in Table.... Columns in
table represent mouse strains used in infection experiments. Rows
represent timepoints post infection plus oocyst and sporozoite
samples. The upper part of the table shows data for first infection,
and oocyst and sporozoite data. The lower part shows data for
challenge infection. Averages were calculated after sample exclusions
(see Methods). For exact values, see Table 1.

	\setlength{\tabcolsep}{14pt}
	\begin{table}[H]
	\small
	\begin{center}
	\caption{Experimental overview with number of read for parasite/mouse indicated.}
\begin{tabular}{*4l}    \toprule
\textit{Day, 1st infection}  	& NMRI  & C57BL/6  & Rag1-/- \\ \midrule
	0 (control)    & $10^4$ / $10^8$  & $10^2$ / $10^8$  & $10^2$ / $10^7$  \\ %rep1/rep2, NA=no replicate
3  		& $10^5$ / $10^8$ & NA  & NA \\ 
5  		& $10^7$ / $10^8$ & $10^5$ / $10^7$  & $10^5$ / $10^7$ \\
7  		& $10^8$ / $10^7$ & NA  & NA \\ 
Oocysts 	& $10^8$ / NA & NA  & NA \\ 
Sporozoites 	& $10^7$ / NA & NA  & NA \\ \midrule

\textit{Day, 2nd infection}  	\\ \midrule
0 (control)     & $10^3$ / $10^8$  &  NA  & NA  \\ %rep1/rep2, NA=no replicate
3  		& $10^4$ / $10^8$ & NA  & NA \\ 
5  		& $10^4$ / $10^8$ & $10^5$ / $10^7$  & $10^5$ / $10^7$ \\
7  		& $10^6$ / $10^8$ & NA  & NA \\

	\bottomrule
 \hline
\end{tabular}
\end{center}
\end{table}

@@

\subsubsection{Alignment and reference genomes}
We used the published \textit{Mus musculus} mm10 assembly (Genome Reference Consortium Mouse 
Build 38, GCA\_000001635.2) as reference genome including annotations for mouse data. The
\textit{E. falciformis} genome (Heitlinger14) was downloaded from ToxoDB (Gajria07). For the
alignment, the mouse and parasite genome files were merged into a dual reference genome, and 
files including mRNA sequences from both species were aligned against the dual reference genome
using TopHat2 (version 2.0.14, Trapnell09) with -G specified, and a Bowtie2 (version 1.1.2, Langmead12) index of the dual genome. Single-end and pair-end sequence samples were aligned separately with library type 'fr-unstranded' specified for pair-end samples. Import into R was enabled by the R package Ballgown, which requires bam files to be processed by Tablemaker (Frazee15), in our case used with -qW -G specified. Tablemaker in turn makes use of Cufflinks (version 2.1.1, Trapnell10).

\subsection{Differential mRNA abundance, data normalisation and sample exclusions}
Complete scripts are available at git@github.com:derele/Ef\_RNAseq.git. Count data was analysed in R, using the upperquartile normalisation method as implemented in edgeR (version 3.14.0). Briefly, genes with zero coverage in all samples (libraries) are removed and normalisation factors are calculated for the 75\% quantile for each library. This  normalisation assumes a negative binomial distribution of read densities. Two samples contradicted the assumption (parasite data) and both mouse and parasite data from these samples were excluded from further analysis: NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 (SI ...). We used a the edgeR pipleine which estimates dispersion values for each gene (provided that a design matrix is defined). The method fits data for each gene to a negative binomial distribution using a generalized linear model (GLM) with the function glmFit() and glmLRT().

........... include outcome of MDS analysis here?

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
A selection of differently abundant mRNAs are used for hierarchical clustering of \textit{E. falciformis} life cycle relevant genes. In each comparison (see Table 3), the union of the 500 differentially abundant mRNAs with lowest FDR are selected. In the next step, the 500 mRNAs from each comparison (or less) are joined. For \textit{E. falciformis} time p.i. comparison, this resulted in 4935 genes of which only the 1618 unique genes were selected. The 22 genes in the NMRI vs C57BL/6 comparison are not included in the \textit{E. falciformis} life cycle analysis. The same comparisons for mouse yielded 8052 genes in total and 1313 unique ones. In heatmaps, all samples, i.e. also samples which themselves did not have any significantly different mRNAs according to our selection, were included in hierarchical clustering. Scale bar in heatmaps show 0 as mean mRNA abundance for each gene (row). Up (green) and down-regulation (brown) denote number of standard deviations from 0, i.e., row mean. Hierarchical clustering was performed using with Euclidean distances, using complete linkage ('complete', R package base).

>>>>

  \section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \section*{Experimental overview}
We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. For each replicate
sample we enriched caecum tissue from three infected mice for
epithelial cells. We extracted mRNA and constructed sequencing
libraries for different timepoints after infection and mouse strains
with assumed different immunocompetence.
%% we have to bring in Simones data first with two sentences and a
%% figure or write it like this
Additonally we obtained data for challenge infections of mice
immunized one month prior to the experiment. For the parasite oocysts
and sporozoites were included as ``environmental'' lifecycle stages.

The number of sequencing reads obtained for individual replicates
ranged between 111111 (samply NMRIxyz) and 222222 (NMRIwyv).

As samples and idividual replicates were sequenced in batches to
different depth and using different instrumentation (see additional
file 1 %%{or include sequencing even in Table 1?
) we performed quality controls confirming the abscence of batch
effects influencing analysis and quality of our results (additional
file xyc).

\section*{Parasite and host transcriptomes can be assessed in parallel}

A maximum of 75\% (sample NMRIaaga) and a minimum of 0.01\% (sample
NMRIagag) of mapped reads could be assigened to the
\textit{E. falciformis} genome in samples considered infected. We
excluded samples NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 due
to an exptremely low parasite contribution to the overall
transcriptome. Technically this made it possible to obtain read counts
following negative bionmial distributions after normalization between
samples (see adddtional file x). Judging from their mRNA output those
samples can be considered non-infected.On the other end of the
spectruem this means that at 7dpi the majority of mRNA originated from
the parasite as opposed to the host transcriptome (Figure 1).

%%%%%
figure 1 label
cotribution of mouse vs. parasite transcripts to the overall
transcriptional output of the analysed tissue for all samples.
%%%%

The total number of sequencing reads obtained as well as those reads
mapped to the transcriptome of either \textit{E. falciformis} and the
mouse are indicated in Table 1 for all replicates.

%% Before was not pretty. All those NA's. If people are like me they
%% hate NAs in tables ;-). I think you sould make the above table from
%% data with xtable.

%% best use a straight 2d table with...

% label & mouse strain & parasite dpi & naive/challenge & number of
% reads & #mapped to mouse & #mapped to \textit{E. falciformis}

%% column name "naive/challenge" can change

%% as technical as it seems people will want to know what our labels
%% mean (they appear in figures throughout) therfore it is good to
%% have them in table 1

\section*{The mouse transcriptome changes in the course of infection infection}

Statistical testing for differences between infected and uninfected
mice, revealed changes in mRNA abundance becoming more pronounced
(involving more genes) at later timepoints of infection (Table 1).

The observed fold-changes were in agreement with previously published
microarry data. Fold-Change data obtained from mice at 6dpi on Agilent
microarrays (Schmidt et al 2012) or 7dpi (our RNAseq data) shows a
strong correlateion (Spearman's \rho = 333; $R^2$ from a linear model
= 444 ; see also figure 2). Considering both and technical differences
between the tow methods and the biological changes along the lifecycle
this confirms the adequacy of our methods for assessing the host
transcriptome.

We selected mouse
%%% This is from methods but all the results should be presented here:

%% Selecting at most at most 500 genes differentially abundant with
%% lowest FDR (<0.05) are selected. In the next step, the 500 mRNAs from
%% each comparison (or less) are joined. For \textit{E. falciformis} time
%% p.i. comparison, this resulted in 1618 unique genes selected as
%% differently expressed . The 22 genes in the NMRI vs C57BL/6 comparison
%% are not included in the \textit{E. falciformis} life cycle
%% analysis. The same comparisons for mouse yielded 8052 genes in total
%% and 1313 unique ones. In heatmaps, all samples, i.e. also samples
%% which themselves did not have any significantly different mRNAs
%% according to our selection, were included in hierarchical
%% clustering.

Those changes are mainly related to the onset of an
adaptive immune response.

\section*{The lifecycle of Eimeria is characterized by large changes in mRNA abundance}

%%% This is from methods but all the results should be presented here:

%% Selecting at most at most 500 genes differentially abundant with
%% lowest FDR (<0.05) are selected. In the next step, the 500 mRNAs from
%% each comparison (or less) are joined. For \textit{E. falciformis} time
%% p.i. comparison, this resulted in 1618 unique genes selected as
%% differently expressed . The 22 genes in the NMRI vs C57BL/6 comparison
%% are not included in the \textit{E. falciformis} life cycle
%% analysis. The same comparisons for mouse yielded 8052 genes in total
%% and 1313 unique ones. In heatmaps, all samples, i.e. also samples
%% which themselves did not have any significantly different mRNAs
%% according to our selection, were included in hierarchical
%% clustering.

%% This might be figure label
%% Scale bar in heatmaps show 0 as mean mRNA abundance for
%% each gene (row). Up (green) and down-regulation (brown) denote number
%% of standard deviations from 0, i.e., row mean.

@@

\subsection{Alignment and reference genomes}
We used the published \textit{Mus musculus} mm10 assembly (Genome
Reference Consortium Mouse Build 38, GCA\_000001635.2) as reference
genome including annotations for mouse data. The
\textit{E. falciformis} genome (Heitlinger14) was downloaded from
ToxoDB (Gajria07). For the alignment, the mouse and parasite genome
files were merged into a dual reference genome, and files including
mRNA sequences from both species were aligned against the dual
reference genome using TopHat2 (version 2.0.14, Trapnell09) with -G
specified, and a Bowtie2 (version 1.1.2, Langmead12) index of the dual
genome. Single-end and pair-end sequence samples were aligned
separately with library type 'fr-unstranded' specified for pair-end
samples. Import into R was enabled by the R package Ballgown, which
requires bam files to be processed by Tablemaker (Frazee15), in our
case used with -qW -G specified. Tablemaker in turn makes use of
Cufflinks (version 2.1.1, Trapnell10).

\subsection{Differential mRNA abundance, data normalisation and sample exclusions}
Count data was normalized using the R-package edgeR (version 3.14.0;
cite) with the upperquartile normalisation method. Briefly, genes with
zero coverage in all samples (libraries) are removed and normalisation
factors are calculated for the 75\% quantile for each library. This
normalisation is suitable for read densities following a negative
binomial distribution. Two samples contradicted this assumption
(parasite data) for later modelling and both mouse and parasite data
from these samples were excluded from further analysis:
NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 (SI ...). The method
then fits a generalized linear model (GLM with a negative binomial
link function) for each gene (glmFit) and then performs likelihood
ratio tests for models w or w/o focal factor (glmLRT).

%% ........... include outcome of MDS analysis here?
%% No, results never are presented in methods.

\subsection{Selection of differentially abundant mRNAs nad hierarchical clustering}
A selection of differently abundant mRNAs are used for hierarchical
clustering of \textit{E. falciformis} life cycle relevant genes. In
each comparison (see Table 3), the union of the at most 500 genes
differentially abundant with lowest FDR (<0.05) are selected. In the
next step, the 500 mRNAs from each comparison (or less) are
joined. For \textit{E. falciformis} time p.i. comparison, this
resulted in 1618 unique genes selected as differently expressed . The
22 genes in the NMRI vs C57BL/6 comparison are not included in the
\textit{E. falciformis} life cycle analysis. The same comparisons for
mouse yielded 8052 genes in total and 1313 unique ones. In heatmaps,
all samples, i.e. also samples which themselves did not have any
significantly different mRNAs according to our selection, were
included in hierarchical clustering. Scale bar in heatmaps show 0 as
mean mRNA abundance for each gene (row). Up (green) and
down-regulation (brown) denote number of standard deviations from 0,
i.e., row mean. Hierarchical clustering was performed using with
Euclidean distances, using complete linkage ('complete', R package
base).

%% the paragraph above has results. All those number

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
All analyses were performed in R (cite R-core). Complete scripts are
available at git@github.com:derele/Ef\_RNAseq.git tagged as version
1.0. 
%% We will make this available as a git webpage probably
